\thetitle

### 2025-05-26 workshop/SPI.tex
NS commnets

\section{Strong parallelism invariance}
\label{sec:PI_causes}
In this section we describe potential violations of 
strong parallelism invariance (SPI) that can occur in a distributed setting. 
We also explain how we avoid these issues by using special 
distributed reduction schemes and splittable random number generators. 
Insights provided in this section can be applied 
to general distributed software beyond Julia.

@@

\subsection{Distributed reduction and floating point non-associativity}
After each round of PT, the machines need to exchange information such as the 
average swap acceptance probabilities, statistics to adapt a variational 
reference and adjust annealing parameters, and so on. 
For instance, suppose our \texttt{state} is univariate and real-valued and 
that each \texttt{Replica} keeps track of the number of times the target chain $N$ 
is visited as well as the mean of the univariate \texttt{state}s from the target chain.
To obtain the final estimate of the mean of the target distribution, we would like 
to pool the mean estimates from each of the \texttt{Replica}s, weighted by 
the number of times that each \texttt{Replica} visited the target chain.   
This process involves summing floating-point values that are located on each 
of the \texttt{Replica}s/machines and is an example of a reduction scheme.

To illustrate why distributed reduction with floating point values can violate 
strong parallelism invariance if not properly implemented, we consider the 
following toy example. 
Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
in the following examples.
In this case, if our reduction procedure is to sum the floating point numbers, 
we know that our final answer should be approximately $36x$. 
However, depending on the exact order in which floating point addition is carried out, 
the answers might not all be the same and exactly equal to $36x$.
For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
for eight machines would be given by 
\[
  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
  \approx 978.5814582452562.
\]
In contrast, with two machines, one possible order of operations might be 
\[
  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
  \approx 978.5814582452563.
\]
A possible reduction tree for two machines is illustrated in \cref{fig:reduction_tree_2}.

@@

@@

A first step to achieve this is to associate one random number 
generator to each \texttt{Replica}. 
To do so, we use our SplittableRandoms.jl package, which is a Julia implementation 
of Java SplittableRandoms. Our package offers an implementation of  
SplitMix64 \cite{steele2014fast},
which allows us to turn one seed into an arbitrary collection of pseudo-independent 
RNGs. 
A quick example of how to use the SplittableRandoms.jl library is given below. 
By splitting a master RNG using the \texttt{split()} function, 
we can achieve SPI even with the use of the \texttt{@threads} macro.

>>>>

\section{Strong parallelism invariance}
\label{sec:PI_causes}
In this section we describe potential violations of SPI that can occur in a 
distributed setting. We also explain how we avoid these issues by using special 
distributed reduction schemes and splittable random number generators. 
Insights provided in this section can be applied 
to general distributed software beyond Julia.

@@

\subsection{Distributed reduction and floating point non-associativity}
The execution of PT in Pigeons.jl is divided into \emph{rounds}, which are
sequences of \emph{scans}---a single call to \cref{alg:distributed_PT}---of 
exponentially increasing length. At the end of each round,
\emph{replicas}---the fundamental unit of parallelization in PT (see \cref{sec:PT} for 
details)---need to exchange information. This can be related to the output of the 
program---e.g., the value of the normalization constant $Z$---as well as statistics 
needed to adapt the sampler so that the next round runs more efficiently.
Almost all of these cases involve summing floating-point numbers that are 
located on different threads, processes, or compute nodes (in the following we use 
the generic term \emph{machine}).

@@

To illustrate why distributed reduction with floating point values can violate 
strong parallelism invariance if not properly implemented, we consider the 
following toy example. 
Suppose we have 8 machines storing the floating point numbers $\cbra{1x, 2x, \ldots, 8x}$, 
as illustrated in \cref{fig:reduction_tree_8}, where we use $x = 10 e^1 \approx 27.1828$
in the following examples.
In this case, if our reduction procedure is to sum the floating point numbers, 
we know that our final answer should be approximately $36x$. 
However, depending on the exact order in which floating point addition is carried out, 
the answers might not all be the same and exactly equal to $36x$.
For instance, in \cref{fig:reduction_tree_8} we see that the order of operations 
for eight machines would be given by 
\[
  ((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x)) \\
  \approx 978.5814582452562.
\]
In contrast, with two machines, one possible order of operations might be 
\[
  (((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x) \\
  \approx 978.5814582452563.
\]
\cref{fig:reduction_tree_2} shows a possible reduction tree for two machines.

@@

A first step to achieve this is to associate one random number 
generator to each replica. 
To do so, we use our SplittableRandoms.jl package, which is a Julia implementation 
of Java SplittableRandoms. Our package offers an implementation of  
SplitMix64 \cite{steele2014fast},
which allows us to turn one seed into an arbitrary collection of pseudo-independent 
RNGs. 
A quick example of how to use the SplittableRandoms.jl library is given below. 
By splitting a master RNG using the \texttt{split()} function, 
we can achieve SPI even with the use of the \texttt{@threads} macro.

### 2025-05-26 workshop/abstract.tex
NS commnets

\begin{abstract}
Sampling from complicated probability distributions---such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics---is a challenging problem. Pigeons.jl provides 
a simple user interface to perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines. In this paper,
we discuss a feature of Pigeons.jl that we call \emph{strong parallelism invariance} (SPI): 
a guarantee that the output for a given seed is identical irrespective of the number 
of threads and processes. This feature is crucial for scientific reproducibility and 
software validation. We describe the key features of Pigeons.jl that enable a 
distributed and randomized algorithm satisfying SPI.
\end{abstract}

>>>>

\begin{abstract}
Sampling from complicated probability distributions---such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics---is a challenging problem. \emph{Pigeons.jl} 
is a Julia package that implements \emph{Parallel Tempering}, a technique that 
leverages parallelism to improve the effectiveness of \emph{Markov chain Monte Carlo} 
algorithms. Pigeons.jl provides a simple user interface to perform such computations 
using single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
machines. In this paper, we discuss a feature of Pigeons.jl that we call \emph{strong 
parallelism invariance} (SPI): a guarantee that the output for a given seed is identical 
irrespective of the number of threads and processes. This feature is crucial for scientific 
reproducibility and software validation. We describe the key features of Pigeons.jl that 
enable a distributed and randomized algorithm satisfying SPI. Finally, we briefly discuss 
some recent developments in Pigeons.jl.
\end{abstract}

### 2025-05-26 workshop/examples.tex
NS commnets

\subsubsection{Online statistics}
\label{sec:online_stats}
Pigeons has facilities to support cases requiring large memory. For instance, we
allow for the computation of online statistics, as well as off-memory sample storage.
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. 
\NS{We should link to the online documentation here and also earlier.}
For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

>>>>

\subsubsection{Online statistics}
\label{sec:online_stats}
Pigeons has facilities to support cases requiring large memory. For instance, we
allow for the computation of online statistics, as well as off-memory sample storage.
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation\footnote{\url{https://pigeons.run/dev/}}.
For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

### 2025-05-26 workshop/introduction.tex
NS commnets

Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible parallel tempering
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1,000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

>>>>

Pigeons.jl\footnote{Documentation available at \url{https://pigeons.run/dev/}. 
Public repository: \url{https://github.com/Julia-Tempering/Pigeons.jl}.}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible \emph{parallel tempering}
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1,000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}
